Implementing Distributed Systems Directly\\with I/O Automata in ioa++

### 2011-09-30 papers/paper1/case_study.tex
Minor tweaks.

Nodes in a distributed system may be arranged in a ring so that two consecutive nodes are connected by a directed channel.
The goal is to designate one of the nodes as the leader of the ring.
The solution we adopt is the asynchronous LCR algorithm presented by Lynch in Chapter 15 of~\cite{lynch1996distributed} which is in turn based on the work of LeLann~\cite{le1977distributed} and Chang and Roberts~\cite{chang1979improved}.
The LCR algorithm assumes that every node has a unique identifier (UID).
When the protocol begins, each node sends its UID to its successor.
When a node receives a UID from its predecessor that is greater than its own, it forwards the UID to its successor.
When a node receives its own UID, it elects itself the leader of the ring.
We add actions for starting an election and withdrawing an election which are necessary to cope with the dynamics discussed previously.
Our implementation of the asynchronous LCR automaton is located in examples/asynch\_lcr\_automaton.hpp\footnote{Access directly at \url{http://github.com/jrwilson/ioa/blob/master/examples/asynch_lcr_automaton.hpp}.} of the ioa++ package.

A node in the ring passively waits for a connection from its successor and actively tries to connect to its successor.
The tcp\_ring\_automaton listed in examples/tcp\_lcr.cpp\footnote{Access directly at \url{http://github.com/jrwilson/ioa/blob/master/examples/tcp_lcr.cpp}.} of the ioa++ package implements this logic and is a composition of an automaton for accepting TCP connections (acceptor), an automaton for making TCP connections (connector), and two automata for TCP connections representing the channel from the predecessor and channel to the successor.
The TCP automata are provided by the ioa++ package and are ordinary automata implemented using file descriptors and the techniques described in Section~\ref{design}.
The tcp\_ring\_automaton uses timeouts and the techniques listed in Section~\ref{constellations} to manage the dynamic constellation of automata and recover from errors.
The tcp\_ring\_automaton contains two external actions:  one for receiving messages from a predecessor and one for sending messages to a successor.

@@

The tcp\_lcr\_automaton which is listed in examples/tcp\_lcr.cpp\footnote{Access directly at \url{http://github.com/jrwilson/ioa/blob/master/examples/tcp_lcr.cpp}.} of the ioa++ package implements the complete protocol by composing an asynch\_lcr\_automaton and a tcp\_ring\_automaton.
Figure~\ref{tcp_lcr_automaton} shows the layered architecture of the TCP LCR system.
The bottom layer consists of component automata for TCP sockets, the middle layer consists of a component automaton that supports coordination via a unidirectional ring, and the top layer implements the business logic.
Thus, ioa++ supports features we expect from a component framework for distributed systems.

>>>>

Nodes in a distributed system may be arranged in a ring so that two consecutive nodes are connected by a directed channel.
The goal is to designate one of the nodes as the leader of the ring.
The solution we adopt is the asynchronous LCR algorithm presented by Lynch in Chapter 15 of~\cite{lynch1996distributed} which is in turn based on the work of LeLann~\cite{le1977distributed} and Chang and Roberts~\cite{chang1979improved}.
The LCR algorithm assumes that every node has a unique identifier (UID).
When the protocol begins, each node sends its UID to its successor.
When a node receives a UID from its predecessor that is greater than its own, it forwards the UID to its successor.
When a node receives its own UID, it elects itself the leader of the ring.
We add actions for starting an election and withdrawing an election which are necessary to cope with the dynamics discussed previously.
Our implementation of the asynchronous LCR automaton is located in examples/asynch\_lcr\_automaton.hpp\footnote{\url{github.com/jrwilson/ioa/blob/master/examples/asynch_lcr_automaton.hpp}} of the ioa++ package.

A node in the ring passively waits for a connection from its successor and actively tries to connect to its successor.
The tcp\_ring\_automaton listed in examples/tcp\_ring\_automaton.hpp\footnote{\url{github.com/jrwilson/ioa/blob/master/examples/tcp_ring_automaton.hpp}} of the ioa++ package implements this logic and is a composition of an automaton for accepting TCP connections (acceptor), an automaton for making TCP connections (connector), and two automata for TCP connections representing the channel from the predecessor and the channel to the successor.
The TCP automata are provided by the ioa++ package and are ordinary automata implemented using file descriptors and the techniques described in Section~\ref{design}.
The tcp\_ring\_automaton uses timeouts and the techniques listed in Section~\ref{constellations} to manage the dynamic constellation of automata and recover from errors.
The tcp\_ring\_automaton contains two external actions:  one for receiving messages from a predecessor and one for sending messages to a successor.

@@

The tcp\_lcr\_automaton which is listed in examples/tcp\_lcr.cpp\footnote{\url{github.com/jrwilson/ioa/blob/master/examples/tcp_lcr.cpp}} of the ioa++ package implements the complete protocol by composing an asynch\_lcr\_automaton and a tcp\_ring\_automaton.
Figure~\ref{tcp_lcr_automaton} shows the layered architecture of the TCP LCR system.
The bottom layer consists of component automata for TCP sockets, the middle layer consists of a component automaton that supports coordination via a unidirectional ring, and the top layer implements the business logic.
Thus, ioa++ supports features we expect from a component framework for distributed systems.

### 2011-09-30 papers/paper1/component_model.tex
refactored abstract and intro, updated my config
Minor tweaks.

In this section we describe how the I/O automata model offers a natural realization of components and interfaces, and provides natural support for concurrency and composition.
%% We then describe extensions to support dynamic semantics as participants in the system may arrive and depart at run-time.
We then describe \textbf{extensions to support dynamic semantics for systems with participants that may arrive and depart at run-time}.

@@

\emph{Hierarchical decomposition}, the modeling dual of recursive encapsulation, is an established technique for I/O automata~\cite{lynch1987hierarchical},~\cite{lynch1994atomic}.
For convenience, we extend the notion that a parent automaton is the composition of its children to include an anonymous child that contains state and actions attributed to the parent.
The \emph{root automaton} is the automaton at the top of the hierarchy and represents the entire system as composition is applied recursively up the hierarchy.

@@

\paragraph*{Binding rules}
To support name-based composition under the I/O automata model, the set of bindings must adhere to four essential rules.
First, the output action and input action of a binding must agree on the type being produced and consumed.
For example, an input action consuming an integer value can only be bound to an output action that produces an integer value.
The system shown in Figure~\ref{sys_model} follows this rule since $(out!(s), in?(s))$ agree on type $s$, $(out!(t), in?(t))$ agree on type $t$, and $(detect!, detected?)$ agree on the absence of a value (i.e., use of a signal).
Second, an input action can be bound to at most one output action.
Composing an input action with multiple output actions would introduce ambiguity into the relationship between the states of the automata and is therefore prohibited by the model.
A binding map where all bindings that mention the same input action are equivalent satisfies this rule.
For dynamics, which are discussed in Section~\ref{dynamics}, we must strengthen this rule to say that an input action can appear at most once in the global binding map.
This strengthened rule means that an input is the sink of at most one binding and that every binding has exactly one owner.
Third, an output action cannot be bound to more than one input action in the same automaton:  the state of the automaton containing the input actions would not be well-defined after the output is executed since the input action effects would be applied in an undefined order.
%% Let $\pi$ be a prefix operator that strips the action name (leaving the path and hence the automaton) of a fully-qualified action name.
%% The following must be true for the global binding map $B$:  $\langle \forall o : (o, z) \in B :: \langle \forall i,j : (o, i) \in B \land (o, j) \in B \land i \neq j:: \pi (i) \neq \pi (j) \rangle \rangle$.
Fourth, an output action cannot be bound to an input action in the same automaton.
Allowing such bindings would result in undefined behavior since the order of the output and input effects is undefined.
The I/O automata model prevents this by requiring that all actions belonging to the same automaton have a unique name.
To elaborate, we usually think about the output action occurring before the input action because a value must be generated.
However, the I/O automata model admits an interpretation where the value is generated first and the effects of the output action and input action are applied non-deterministically.
Graphically, an arrow originating in one component must terminate in another component, \textbf{even if the latter is a child of the former} as Figure~\ref{sys_model} illustrates.
%% To adhere to this rule, the following must be true for the global binding map $B$:  $\langle \forall o, i : (o, i) \in B :: \pi (o) \neq \pi (i) \rangle$.

@@

While assuming a fixed number of automata is reasonable for modeling, assuming a fixed number of automata for a real system may result in either over-provisioning or inflexibility.
To develop a component based on the static I/O automata model, a developer must choose a concrete value for the abstract quantity $N$.
System resources are wasted if the actual number of participants is much smaller than $N$.
If the number of participants is greater than $N$, then the system cannot respond to certain situations even though resources might be available.
Thus, the techniques for \emph{modeling} a dynamic system of I/O automata are not necessarily appropriate for \emph{implementing} a dynamic system of I/O automata \textbf{directly}.
%% Consequently, we add the ability to deal with a dynamic set of automata and interactions.

>>>>

In this section we describe how the I/O automata model offers a natural realization of components and interfaces, and provides natural support for concurrency and composition.
%% We then describe extensions to support dynamic semantics as participants in the system may arrive and depart at run-time.
We then describe extensions to support semantics for dynamic systems, i.e., systems whose set of constituent automata changes over time.

@@

%The I/O automata formalism was developed to model reactive systems~\cite{lynch1987hierarchical} and has been used in the design and verification of concurrent and distributed systems.
I/O automata are a good foundation for such components because they have independent state, straightforward interfaces, and precise semantics for event generation, distribution, and consumption resulting in well-defined interactions under composition.
\emph{Hierarchical decomposition}, the modeling dual of recursive encapsulation, is an established technique for I/O automata~\cite{lynch1987hierarchical},~\cite{lynch1994atomic}.
For convenience, we extend the notion that a parent automaton is the composition of its children to include an anonymous child that contains state and actions attributed to the parent.
The \emph{root automaton} is the automaton at the top of the hierarchy and represents the entire system as composition is applied recursively up the hierarchy.

@@

\paragraph*{Binding rules}
To support name-based composition under the I/O automata model, the set of bindings must adhere to four essential rules.
First, the output action and input action of a binding must agree on the type being produced and consumed.
For example, an input action consuming an integer value can only be bound to an output action that produces an integer value.
The system shown in Figure~\ref{sys_model} follows this rule since $(out!(s), in?(s))$ agree on type $s$, $(out!(t), in?(t))$ agree on type $t$, and $(detect!, detected?)$ agree on the absence of a value (i.e., use of a signal).
Second, an input action can be bound to at most one output action.
Composing an input action with multiple output actions would introduce ambiguity into the relationship between the states of the automata and is therefore prohibited by the model.
A binding map where all bindings that mention the same input action are equivalent satisfies this rule.
For dynamics, which are discussed in Section~\ref{dynamics}, we must strengthen this rule to say that an input action can appear at most once in the global binding map.
This strengthened rule means that an input is the sink of at most one binding and that every binding has exactly one owner.
Third, an output action cannot be bound to more than one input action in the same automaton:  the state of the automaton containing the input actions would not be well-defined after the output is executed since the input action effects would be applied in an undefined order.
%% Let $\pi$ be a prefix operator that strips the action name (leaving the path and hence the automaton) of a fully-qualified action name.
%% The following must be true for the global binding map $B$:  $\langle \forall o : (o, z) \in B :: \langle \forall i,j : (o, i) \in B \land (o, j) \in B \land i \neq j:: \pi (i) \neq \pi (j) \rangle \rangle$.
Fourth, an output action cannot be bound to an input action in the same automaton.
Allowing such bindings would result in undefined behavior since the order of the output and input effects is undefined.
The I/O automata model prevents this by requiring that all actions belonging to the same automaton have a unique name.
To elaborate, we usually think about the output action occurring before the input action because a value must be generated.
However, the I/O automata model admits an interpretation where the value is generated first and the effects of the output action and input action are applied non-deterministically.
Graphically, an arrow originating in one component must terminate in another component as Figure~\ref{sys_model} illustrates.
%% To adhere to this rule, the following must be true for the global binding map $B$:  $\langle \forall o, i : (o, i) \in B :: \pi (o) \neq \pi (i) \rangle$.

@@

While assuming a fixed number of automata is reasonable for modeling, assuming a fixed number of automata for a real system may result in either over-provisioning or inflexibility.
To develop a component based on the static I/O automata model, a developer must choose a concrete value for the abstract quantity $N$.
System resources are wasted if the actual number of participants is much smaller than $N$.
If the number of participants is greater than $N$, then the system cannot respond to certain situations even though resources might be available.
Thus, the techniques for \emph{modeling} a dynamic system of I/O automata are not necessarily appropriate for \emph{implementing} a dynamic system of I/O automata.
%% Consequently, we add the ability to deal with a dynamic set of automata and interactions.

### 2011-09-30 papers/paper1/design.tex
Minor tweaks.

A more focused approach used in our framework is to allow automata to declare \emph{dynamically} the actions they would like the scheduler to consider.
The controller then has the responsibility of processing the set of requested actions according to some fair policy.
Let an action be \emph{enabled} if its precondition is true and \emph{disabled} otherwise.
Observe that executing an action potentially changes the state of the automaton and causes the set of enabled actions to change.
Consequently, an automaton is given the opportunity to declare actions to the scheduler after one of its actions is executed.
Since the state variables (and therefore preconditions) of all automata are independent, each automaton is responsible for scheduling its own actions.
For bootstrapping, each automaton is allowed to schedule actions when it is created.

@@

However, an automaton should still ensure that the scheduler will \emph{consider} all actions that are currently enabled.
A simple technique for accomplishing this is to \emph{schedule} all enabled actions.
More efficient schemes are possible if one considers how an action might enable other actions and which actions have already been scheduled.
However, this must be done carefully.
Not declaring an enabled action would violate the fairness of the controller since it will never select an action that may cause a state change.
Not scheduling an (enabled) action \textbf{could be another potential mistake}.

>>>>

A more focused approach used in our framework is to allow automata to declare \emph{dynamically} the actions they would like the scheduler to consider.
The act of declaring an action to the scheduler is called \emph{scheduling}.
The controller then has the responsibility of processing the set of requested actions according to some fair policy.
Let an action be \emph{enabled} if its precondition is true and \emph{disabled} otherwise.
Observe that executing an action potentially changes the state of the automaton and causes the set of enabled actions to change.
Consequently, an automaton is given the opportunity to schedule actions to the scheduler after one of its actions is executed.
Since the state variables (and therefore preconditions) of all automata are independent, each automaton is responsible for scheduling its own actions.
For bootstrapping, each automaton is allowed to schedule actions when it is created.

@@

However, an automaton should still ensure that the scheduler will \emph{consider} all actions that are currently enabled.
A simple technique for accomplishing this is to schedule all enabled actions.
More efficient schemes are possible if one considers how an action might enable other actions and which actions have already been scheduled.
However, this must be done carefully.
Not scheduling an enabled action would violate the fairness of the controller since it will never select an action that may cause a state change.
Not scheduling an (enabled) action is a common mistake.

### 2011-09-30 papers/paper1/evaluation.tex
Minor tweaks.

The driver for a single trial can be found in examples/random.cpp\footnote{Access directly at \url{http://github.com/jrwilson/ioa/blob/master/examples/random.cpp}.} of the ioa++ package.
The trials were performed on a Mac G4 running OS X 10.5.8 with two 2.66 GHz Dual-Core Intel Xeon processors with 2GB of memory.
The code was compiled and linked with i686-apple-darwin9-g++-4.0.1 build 5493 with -O2 optimization and the -DPROFILE flag to enable profiling in ioa++.

>>>>

The driver for a single trial can be found in examples/random.cpp\footnote{\url{github.com/jrwilson/ioa/blob/master/examples/random.cpp}} of the ioa++ package.
The trials were performed on a Mac G4 running OS X 10.5.8 with two 2.66 GHz Dual-Core Intel Xeon processors with 2GB of memory.
The code was compiled and linked with i686-apple-darwin9-g++-4.0.1 build 5493 with -O2 optimization and the -DPROFILE flag to enable profiling in ioa++.
